Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-23 icfp.tex
Pass over till end of 3.1

In this section, we develop a linear calculus $\lambda_\Delta^\pi$,
dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
and multiplicity polymorphism introduced by Linear
Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
language (algebraic datatypes, case expressions, and recursive lazy let bindings)
except for equality coercions, which we discuss as future work in
Section~\ref{sec:future-work}.
%
%
Linear Core makes precise the various insights discussed in the
previous section by crystallizing them together in a linear type system for which we
prove soundness via the standard type preservation and progress theorems. Crucially,
Linear Core typing accepts all the \emph{semantically linear} examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Besides type safety, we prove that multiple optimising Core-to-Core
transformations preserve linearity in Linear Core, while violating linearity
under Core's current type system.
%
% As far as we know, we are the first to prove optimisations preserve types in
% a non-strict linear language.

The first key idea for typing linearity semantically is to delay \emph{consuming a
resource} to when a computation that depends on that resource is effectively
evaluated or returned, by annotating relevant binders with \emph{usage
environments} (Section~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for case
expressions, branching on whether the scrutinee is in Weak Head Normal
Form, and using ``proof irrelevance'' to track resources that are no
longer available but haven't yet been fully consumed
(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
resources between the pattern-bound variables, %^as usage environments
encoding that all linear components of a pattern must be used.
% 
We also note that despite the focus on GHC Core, the fundamental ideas for
understanding linearity in a call-by-need calculus can be readily applied to
other non-strict languages.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to typecheck case expressions and alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D_1}\s_1\dots x_n{:}_{\D_n}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for type-checking semantic linearity in the presence of laziness we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default} or \emph{wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting whether the pattern-bound variables must
consumed linearly (ultimately, in order to consume the scrutinee linearly).
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.

>>>>

In this section, we develop a linear calculus $\lambda_\Delta^\pi$,
dubbed \emph{Linear Core}, that combines the linearity-in-the-arrow
and multiplicity polymorphism introduced by Linear
Haskell~\cite{cite:linearhaskell} with all the key features from GHC's Core
language (algebraic datatypes, case expressions, and recursive lazy let bindings),
except for equality coercions, which we discuss as future work in
Section~\ref{sec:future-work}.
%

Linear Core makes precise the various insights discussed in the
previous section in a linear type system that we prove is sound and guarantees
linear resource usage at runtime via a standard soundness argument on a
call-by-need big-step operational semantics that becomes stuck when linear
variables in the runtime heap are used more than once.
\todo{O que digo aqui sobre a semantica launchbury style?}
Crucially,
Linear Core typing accepts all the semantically linear examples
(highlighted with \colorbox{notyet}{\notyetcolorname})
from Section~\ref{sec:semantic-linearity-examples}, which Core currently
rejects.
%
Additionally, we prove that multiple optimising Core-to-Core transformations
preserve linearity in Linear Core, while violating it under Core's
current type system.
%
%To the best of our knowledge, we are the first to prove optimisations preserve
%types in a non-strict linear language.
% 
Despite the focus on GHC Core, the fundamental ideas for
understanding linearity in our call-by-need calculus can be readily applied to
other non-strict languages.

Linear Core's first key idea is to delay consuming
resources until a computation that depends on those resources is effectively
evaluated or returned. Delayed consumption is tracked by annotating relevant
binders with \emph{usage environments} (Section~\ref{sec:usage-environments}).
%
The second key idea is to have two distinct rules for case
expressions, branching on whether the scrutinee is in Weak Head Normal
Form or not, and using ``proof irrelevance'' to track resources that are no
longer available but haven't yet been fully consumed
(Section~\ref{sec:lc-case-exps}). Additionally, we introduce tags to split
resources amongst pattern-bound variables and %^as usage environments
encode that all linear components of a pattern must be used together.

We present Linear Core's syntax and type system incrementally, starting with the
judgements and base linear calculi rules (Section~\ref{sec:base-calculi}).
%
Then, usage environments, the rule for $\Delta$-bound variables, and rules for
(recursive) let bindings (Section~\ref{sec:usage-environments}).
%
Finally, we introduce the rules to type case expressions and case alternatives,
along with the key insights to do so, namely branching on WHNF-ness of
scrutinee, proof irrelevant resources, and tagging (Section~\ref{sec:lc-case-exps}).

@@

%
Terms are variables $x,y,z$, data constructors $K$, multiplicity
abstractions $\Lambda p.~e$ and applications $e~\pi$, term abstractions
$\lambda \x[\pi].~e$ and applications $e~e'$, where lambda binders are
annotated with a multiplicity $\pi$ and a type $\s$. Then, there are
non-recursive let bindings $\llet{\xD = e}{e'}$, recursive let bindings
$\lletrec{\ov{\xD = e}}{e'}$, where the overline denotes a set of distinct
bindings $x_1{:}_{\D}\s_1\dots x_n{:}_{\D}\s_n$ and associated expressions
$e_1\dots e_n$, and case expressions $\ccase{e}{\zD~\{\ov{\rho \to e'}\}}$,
where $z$ is the case binder and the overline denotes a set of distinct
patterns $\rho_1\dots \rho_n$ and corresponding right hand sides $e'_1\dots
e'_n$. We often omit the case binder $z$ when it is unused. Notably, (recursive) let-bound
binders and case-bound binders are annotated
with a so-called \emph{usage environment} $\Delta$ -- a fundamental construct
for typing semantic linearity in the presence of laziness that we present in
Section~\ref{sec:usage-environments}.
%
Case patterns $\rho$ can be either the \emph{default/wildcard}
pattern $\_$, which matches any expression, or a constructor $K$ and a set of
variables that bind its arguments, where each field of the constructor has an
associated multiplicity denoting which pattern-bound variables must be
consumed linearly or unrestrictedly.
Additionally, the set of patterns in a case expression is assumed to be exhaustive,
i.e. there is always at least one pattern which matches the scrutinized expression.
